One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-26 paper/model.tex
minor updates
second draft

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

@@

@@

\subsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which circumvents time-step
constraint set by the inner disc boundary \citep{masset00a,masset00b}.

@@

% poisson kernal integration, softening 
% where used?
% BC
\subsection{ZEUS-MP}
ZEUS-MP  is a general-purpose finite difference
code \citep{hayes06}. We use the code in 3D spherical geometry, covering
$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
The grid is logarithmically spaced in radius and uniformly spaced in the angular
coordinates. We assume reflectional symmetry across the midplane, and
apply reflective boundary conditions at radial boundaries and the
upper disc boundary.

@@

\subsubsection{Evolution of non-axisymmetirc modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
  &= f_0(R) + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

The normalised surface density with azimuthal wavenumber $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
the real frequency and $\gamma$ is the growth rate. The total
non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

\subsubsection{Angular momentum decomposition}
The total disc angular momentum is
\begin{align}
  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR 
  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
  %dR 
  = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation. We will refer to $J_m$ as the
$m^\mathrm{th}$ component of the total angular momentum, and use it to
monitor angular momentum conservation in the numerical simulations. 
It is important to distinguish this empirical definition from the
angular momentum of linear perturbations, defined through a
conservation law, which is given in \S\ref{wkb}.

@@

\section{Linear density waves}\label{wkb}
In this paper we take a fully numerical approach using non-linear
hydrodynamic simulations. However, it will be useful to employ
linear density wave theory to interpret the simulations.
We list here the main definitions and results from 2D linear theory,
taken from \cite{lin11b} and \cite{shu91}.

In a linear analysis, one assumes a steady axisymmetric background state, such as
that defined in \S\ref{model}, which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables. We take $m>0$. Since the background is
axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
    t\right)} = \Sigma_m$ as defined for diagnostics. 
The linearised mass and momentum equations are
\begin{align}
  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\\
  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
\end{align}
where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicyclic frequency. A
locally isothermal equation of state has been assumed.    
The linearised Poisson equation is 
\begin{align}
  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
\end{align}

The linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
exploration of the linear problem to a future study. Here, we list some
general  properties of the linear perturbations that will be useful
for describing simulation results.

\subsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagrangian
displacement, and $\bm{F}$ is the vertically-integrated angular
momentum flux consisting of a Reynolds stress and a gravitational
stress \citep{lin11b}. Explicit expressions for $\bm{\xi}$ can be found in,
e.g. \cite{papaloizou85}. 
%and $\bm{F}$ can be read off Eq. \ref{angflux1}.

% \cite{papaloizou85} and \cite{lin11} respectively, and

% lin papaloizou kley 1993, ryu/goodman, narayan

The background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the total angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momentum flux.% and the perturbations are regular.

However, as noted in \cite{lin11b}, if there is an imposed
temperature gradient, as in the disc models we consider,
then $\tbg\neq 0$ in general, which corresponds to a torque
exerted by the background disc on the perturbation. We 
outline the derivation of $T_\mathrm{BG}$ in Appendix 
\ref{tbg_deriv}.

\subsection{Local results}
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. The
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc is  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
Note that in this approximation, only axisymmetric perturbations with
$m=0$ can be unstable (having an imaginary $\sigma$).

>>>>

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
 disc with constant $H/R$.

@@

\section{Linear density waves}\label{wkb}
In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
stability of disc equilibra described 
above. However, it is useful to  interpret simulation results
using linear density theory. We list here the main definitions and
results from 2D linear theory, taken from \cite{lin11b} and
\cite{shu91}.   In particular, we highlight a key feature of
locally isothermal discs that enables angular momentum exchange
between linear perturbations and the background disc.

In a linear analysis, one assumes a steady axisymmetric background state, such as
that defined in \S\ref{model}, which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
loss of generality.  
%Since the background is
%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
 %   t\right)} = \Sigma_m$ as defined for diagnostics.

The linearised mass and momentum equations are
\begin{align}
  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\label{radmom}\\
  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
\end{align}
where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicyclic frequency. A
locally isothermal equation of state has been assumed in
Eq. \ref{radmom}.  The linearised Poisson equation is 
\begin{align}
  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
\end{align}

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
exploration of the linear problem to a future study. Here, we list some
general  properties of the linear perturbations that will be relevant
for describing simulation results.

\subsection{Global angular momentum conservation for linear
  perturbations} \label{global_cons}
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align}\label{angcons}
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density of the linear disturbance (which may
be positive or negative), where $\bm{\xi}$ is the Lagrangian
displacement and $^*$ denotes complex conjugation, and $\bm{F}$ is the
vertically-integrated angular momentum flux consisting of a Reynolds
stress and a gravitational stress \citep{lin11b}. Explicit expressions
for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85}.

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the total angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momentum flux. % and the perturbations are regular.

However, as noted in \cite{lin11b}, if there is an imposed
temperature gradient, as in the disc models we consider,
then $\tbg\neq 0$ in general, which corresponds to a local torque
exerted by the background disc on the perturbation. We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.

The important consequence of the background torque is the possibilty
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is positive (negative),
then the local angular momentum density of the linear disturbance
further increases (decreases). This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc. In our disc models with $c_s^2\propto R^{-q}$, we expect the
growth rate of such an instability to scale with $q$.

\subsection{Local results}
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. The
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc is  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
Note that in the local approximation, only axisymmetric perturbations
($m=0$) can be unstable. % (having an imaginary $\sigma$).

@@

\subsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which removes the mean azimuthal
velocity of the disc, thereby permit larger timesteps than that would otherwise be allowed
by the usual Courant condition based on the full azimuthal
velocity \citep{masset00a,masset00b}.

@@

% poisson kernal integration, softening 
% where used?
% BC
\subsection{ZEUS-MP}
ZEUS-MP  is a general-purpose finite difference
code \citep{hayes06}. We use the code in 3D spherical geometry, covering
$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
The grid is logarithmically spaced in radius and uniformly spaced in the angular
coordinates. We assume symmetry across the midplane, and
apply reflective boundary conditions at radial boundaries and the
upper disc boundary.

@@

\subsubsection{Evolution of non-axisymmetirc modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

The normalised surface density with azimuthal wavenumber $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$ as in linear
theory. 
%where
%$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
%the real frequency and $\gamma$ is the growth rate. 
The total non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

\subsubsection{Angular momentum decomposition}
The total disc angular momentum is
\begin{align}
  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR 
  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
  %dR 
  = \sum_{m=0}^\infty J_m. 
\end{align}
We will refer to $J_m$ as the
$m^\mathrm{th}$ component of the total angular momentum, and use it to
monitor numerical angular momentum conservation in the simulations. 
It is important to distinguish this empirical definition from the
angular momentum of linear perturbations given in \S\ref{wkb}, which
is defined through a conservation law.

@@
